\Large{All in the family:\\Anna and Lisa Hahner's finishing times in\\the 2016 women's Olympic marathon}

### 2017-02-02 paper/paper.tex
paper updates

On August 14, 2016, at 9:30 in the morning, the Women's Olympic
marathon kicked off in Rio de Janeiro, Brazil, when 156 runners from
80 countries across the world left the starting line en route to their
destination 42.195 kilometers away. Two hours, twenty-four minutes,
and four seconds later, Jemima Sumgong of Kenya would be the first to
cross the finishline and take home gold; Sumgong was just 3.5 minutes
behind her personal best time in the marathon. Approximately 21
minutes later, twin marathoners from Germany, Anna and Lisa Hahner,
would cross the finishline together, holding hands and celebrating a
personal victory. Although the Hahners would finish 81st and 82nd,
respectively, with times slightly more than 18 minutes slower than
corresponding personal bests, Anna Hahner would describe their joint
finish as a ``magical moment.''
\href{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}
The media quickly picked up on the Hanher story as an image of the
beaming twins finishing hand-in-hand captured a public audience. Many
believed the moment was a reflection of the Olympic spirit.

Not everyone agreed with this rosy interpretation. The twins' happy
facial expressions at the finish were portrayed as a bit contrived
(smiling like ``Honigkuchenpferde''---a Honigkuchenpferd is a German
cookie in the shape of a horse), and the sports director of the German
Athletics Federation, Thomas Kurschilgen, stirred up controversy when
he suggested that the Hahners' photo-finish was no coincidence.
Kurschilgen averred that the twins slowed down so as to finish
simultaneously and create a spectacle which would ``generate media
attention.'' Kurschilgen justified his charge with the fact that the
twins ran in the Rio marathon at least 18 minutes slower their
personal best times prior to the
Olympics.\footnote{\url{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}}
Not surprisingly, these accusations were denied by the Hahner twins,
who claimed that their simultaneous finish was simply an unintended
coincidence.

@@

What happened in the women's Olympic marathon, and how might we
develop a statistical approach that assesses whether the Hahner finish
was coincidental or intentional?  These two interpretations are
clearly at odds. If the former, then the Hahner twins are to be
celebrated and their finish treated as an expression of the spirit
behind the Olympic games. If the latter, though, then the twins may
have violated this spirt by not trying hard enough. It is perhaps too
easy for us to write such a glib sentence---neither of us cannot
fathom being able to complete a marathon anywhere in the vicinity of
two and a half hours---but we nonetheless want to know what the data
from the Olympic marathon data tell us.  Was the Hahner finish in the
2016 women's Olympic marathon a lovely coincidence or something else?

Among female Olympic marathoners, the Hahner twins were not alone in
their familial ties, and we draw on this in the analysis that follows.
The marathon also featured twins from North Korea, Kim Hye-song and
Kim Hye-gyong, who posted identical times and finished 10th and 11th
in the race, respectively. The Kim finish, unlike the Hahner finish,
appears devoid of post-race controversy. Moreover, three triplets from
Estonia competed in the Rio marathon, although only two, Lily Luik and
Leila Luik, finished it, in 97th and 114th place, respectively. The
third Estonia triplet, Liina Luik, recorded what is known as a
DNF---an abbreviation that means did not finish, a term that we will
use throughout this article.

@@

For each participant who started the women's Olympic marathon, we know
several things: personal best marathon time prior to the 2016 Olympic
games; split times from the Rio marathon course at 5 kilometers, 10
kilometers, and so forth; and, finishing time. We cannot directly
observe the effort that an individual put into the race, and we do not
know why some runners have DNF results.  Some runners may have injured
themselves on the course and accordingly dropped out, and others may
have dropped out, uninjured, in anticipation of an unsatisfactory
result. Of the 156 marathon starters, 133 completed the race and 23
DNFed at various locations throughout the course. The overall DNF rate
was thus $\frac{23}{156} \approx 0.15$, and the relatively small
sample size at our disposal means that a 95\% confidence for this rate
is fairly wide, namely, $\left(0.098, 0.22\right)$.  Key in the
analysis that follows is the deviation between a runner's Rio time and
her prior personal best time in the marathon.  While there is variance
across runners in these deviations, they nonetheless may be able to
provide us with insight as to the Hahner twins' performance in the
marathon.

Our research design has two components.  In the first, we present some
visualizations that present the Hahner twin's results in context; our
interest here is primarily the two Hanhers although we comment on the
Kim twins and Luik triplets when appropriate.  Our visualizations
suggest blah blah bhah.  We then turn to a modeling approach and in
particular a simulation.  In simulated Olympic marathons, which we
develop based on relationships (excluding twins and triplets) between
known personal best times and observed Rio finishing times, we find
that the Hahner twins finished suspiciously close to each other given
the disparty between their personal best times.  OR THEY DID NOT.

@@

SOME PLOTS HERE.  Maybe PB versus FINAL and HALF vs Final.

@@

\begin{figure}[!ht]
  \caption{Relationship between Personal Best and Result  -- caption?}
  \label{fig:scatter}
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[width=\textwidth,
    keepaspectratio]{scatter_plot.pdf}
    \caption{Rio finishing times and personal best times}
    \label{fig:45degreeplot}
  \end{subfigure}
  \begin{subfigure}{.5\textwidth}
    \centering
    % \label{fig:Studentized Residuals}
    \includegraphics[width=.975\textwidth, keepaspectratio]{studentized_residuals.pdf}
    \caption{Studentized residuals}
    \label{fig:studentizedresiduals}
  \end{subfigure}
\end{figure}

@@

\begin{figure}[!ht]
 \caption{Relationship between Difference in Personal Best and Difference in Result}
 \label{fig:diffdiffscatter}
 \centering
 \begin{subfigure}{.5\textwidth}
 \includegraphics[width=\textwidth, keepaspectratio]{diff_in_diff_scatter_plot.pdf}
 \end{subfigure}
\end{figure}

>>>>

On August 14, 2016, at 9:30 in the morning, the Women's Olympic
marathon kicked off in Rio de Janeiro, Brazil, when 156 runners from
80 countries across the world left the starting line en route to their
destination 42.195 kilometers away. Two hours, twenty-four minutes,
and four seconds later, Jemima Sumgong of Kenya would be the first to
cross the finishline and take home gold; Sumgong was just 3.5 minutes
behind her prior personal best time in the marathon. Approximately 21
minutes later, twin marathoners from Germany, Anna and Lisa Hahner,
would cross the finishline together, holding hands and celebrating a
personal victory. Although the Hahners would finish 81st and 82nd,
respectively, with times slightly more than 18 minutes slower than
corresponding personal bests, Anna Hahner would describe their joint
finish as a ``magical moment.''
\href{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}
The media quickly picked up on the Hanher story as an image of the
beaming twins finishing hand-in-hand captured a public audience. Many
believed the moment was a reflection of the Olympic spirit.

Not everyone agreed with this rosy interpretation. The twins' happy
facial expressions at the finish were portrayed as a bit contrived
(smiling like ``Honigkuchenpferde''---a Honigkuchenpferd is a German
cookie in the shape of a horse), and the sports director of the German
Athletics Federation, Thomas Kurschilgen, stirred up controversy when
he suggested that the Hahners' photo-finish was no coincidence.
Kurschilgen averred that the twins slowed down so as to finish
simultaneously and create a spectacle which would ``generate media
attention.'' Kurschilgen justified his charge with the fact that the
twins ran in the Rio marathon at least 18 minutes slower their
personal best times prior to the
Olympics.\footnote{\url{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}}
Not surprisingly, Kurschilgen's accusations were denied by the Hahner
twins, who claimed that their simultaneous finish was simply an
unintended coincidence.

@@

What happened in the women's Olympic marathon, and how might we
develop a statistical approach that assesses whether the Hahner twin's
finish was coincidental or intentional?  These two interpretations are
clearly at odds. If the former, then the Hahners are to be celebrated
and their finish treated as an expression of the spirit behind the
Olympic games. If the latter, though, then the twins may have violated
this spirt by not trying hard enough. It is perhaps too easy for us to
write such a glib sentence---neither of us can fathom being able to
complete a marathon anywhere in the vicinity of two and a half
hours---but we nonetheless want to know what the data from the Olympic
marathon data tell us.  Was the Hahner finish in the 2016 women's
Olympic marathon a lovely coincidence or something else?

Among female Olympic marathoners, the Hahner twins were not alone in
their familial ties, and we touch on this in the analysis that
follows.  The marathon also featured twins from North Korea, Kim
Hye-song and Kim Hye-gyong, who posted identical times and finished
10th and 11th in the race, respectively. The Kim finish, unlike the
Hahner finish, appears devoid of post-race controversy. Moreover,
three triplets from Estonia competed in the Rio marathon, although
only two, Lily Luik and Leila Luik, finished it, in 97th and 114th
place, respectively. The third Estonia triplet, Liina Luik, recorded
what is known as a DNF---an abbreviation that means did not finish, a
term that we will use throughout this article.

@@

For each participant who started the women's Olympic marathon, we know
several things: personal best marathon time prior to the 2016 Olympic
games; split times from the Rio marathon course at 5 kilometers, 10
kilometers, and so forth; and, finishing time. We cannot directly
observe the effort that an individual put into the race, and we do not
know why some runners have DNF results.  Some runners may have injured
themselves on the course and accordingly dropped out, and others may
have dropped out, uninjured, in anticipation of an unsatisfactory
result. Of the 156 marathon starters, 133 completed the race and 23
DNFed at various locations throughout the course. The overall DNF rate
was thus $\frac{23}{156} \approx 0.15$, and the relatively small
sample size at our disposal means that a 95\% confidence for this rate
is fairly wide, namely, $\left(0.098, 0.22\right)$.

The Kurschilgen accusation against the Hahner twins has two
components, that the twins finished simultaneously, holding hands and
smiling, \emph{and} that they ran slowly.  We suspect that Kurschilgen
would not have expressed ire at the Hahners had they finished in 1st
and 2nd place in Rio, hand-in-hand with wide grins, but of course we
do not know this.  As such, our investigation of the charges that
Kurschilgen offered will distinguish between the idea of slow finish
versus a simultaneous finish.

Our research design has two components.  In the first, we present
visualizations that describe various features of the Hahner twin's
results, and an important element of our visualizations is the
difference between a runner's Rio time and her prior personal best
time in the marathon.  Overall, our visualizations suggest blah blah
bhah.  We then turn to modeling and in particular a regression-based
simulation.  In simulated Olympic marathons, which we develop based on
relationships (excluding the Hahner and Kim twins and Luik triplets)
between known personal best times and observed Rio finishing times, we
find that the Hahner twins finished suspiciously close to each other
given the disparty between their personal best times.  OR THEY DID
NOT.  Finally, we return to Kurschilgen claims  about the Hahners and
offer our thoughts about their validity.

@@

Figure \ref{fig:scatter} contains two plots, both of which describe
how the Rio Olympic women's marathon results varied as a function of
athletes' personal best times and their half marathon split times.
The points in each plot are colored by twin/triplet status, and both
plots contain second-order polynomial regression lines.  We treat an
athlete's prior personal best marathon time as a measure of the
athlete's underlying marathon talent, and we interpret an athlete's
half marathon split as a measure of the athlete's skill at marathon
running on August 14, 2016, the date of the women's marathon in Rio.

\begin{figure}[!ht]
  \caption{Olympic results and measures of marathon ability}
  \label{fig:scatter}
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[width=\textwidth, keepaspectratio]{scatter_plot.pdf}
    \caption{Finishing and personal best times}
    \label{fig:45degreeplot}
  \end{subfigure}
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[width=\textwidth, keepaspectratio]{scatter_plot_half.pdf}
    \caption{Finishing and half marathon split times}
    \label{fig:45degreeplot_half}
  \end{subfigure}
\end{figure}

Considering first Figure \ref{fig:45degreeplot_half}, note that this
figure includes a solid 45-degree line as the horizontal and vertical
axes in the figure correspond to full marathons.  The rug marks on the
bottom of the figure denote the personal best times of runners who
DNFed.  Given the paucity of points (five of them) below the 45-degree
line, the vast majority of Olympic marathoners ran slower in Rio
compared to their personal bests.  Relative to personal best times,
the Hahner twins were definitely on the slow side, but a number of
runners had greater differences between their Olympic times and their
personal bests than the Hahners.  These runners are denoted with
squares in Figure \ref{fig:45degreeplot}, and there are 13 such
symbols, highlighting approximately 10\% of the racers who completed
the marathon.  The figure shows moreover that there were two women who
had personal best times slightly faster than the Hahner's and yet
finished after both of the German women.  Although Figure
\ref{fig:45degreeplot} suggests that the Hahner twins were slower than
one would have expected given their previous best marathon times, it
is not consistent with the accusation that they dramatically slowed
down in the Rio marathon.

% > df <- read.csv("data/times.csv", stringsAsFactors = FALSE) %>% tbl_df()
% > df <- df[df$BIB != 1172,]
% > sort(dta$FINAL - dta$PB) / 60
%   [1] -2.3333333 -2.2166667 -2.0666667 -1.1166667 -0.4000000  0.2166667
%   [7]  0.6333333  0.9000000  1.0166667  1.3666667  1.5166667  1.7166667
%  [13]  1.7666667  1.8500000  2.1166667  2.4666667  2.5333333  3.0166667
%  [19]  3.0500000  3.1666667  3.2666667  3.2833333  3.3666667  3.4500000
%  [25]  3.6833333  3.9000000  3.9833333  4.2000000  4.2666667  4.3833333
%  [31]  4.4666667  4.5833333  4.6166667  4.6333333  4.6833333  4.7333333
%  [37]  4.7333333  4.8333333  5.0666667  5.0833333  5.0833333  5.1000000
%  [43]  5.1333333  5.3500000  5.3500000  5.4000000  5.7166667  5.7333333
%  [49]  5.9333333  6.0000000  6.0333333  6.1166667  6.3333333  6.3500000
%  [55]  6.5000000  6.5333333  6.6000000  6.6666667  6.7166667  6.7833333
%  [61]  6.8166667  6.9500000  7.1000000  7.1500000  7.2000000  7.3500000
%  [67]  7.4500000  7.5666667  7.6000000  7.6000000  7.6333333  7.7500000
%  [73]  7.8833333  7.9666667  7.9833333  8.3000000  8.6000000  9.3000000
%  [79]  9.5500000  9.5833333  9.6000000  9.7166667  9.9500000  9.9833333
%  [85] 10.3166667 10.5333333 10.6666667 10.7000000 10.8333333 10.9666667
%  [91] 11.1500000 11.1833333 11.4666667 11.7166667 12.1666667 12.6000000
%  [97] 12.9166667 13.2166667 13.2833333 13.3833333 13.7666667 13.8666667
% [103] 13.9166667 14.0500000 14.0833333 14.5666667 14.6833333 14.7500000
% [109] 15.0000000 15.0500000 15.5000000 15.9000000 16.2000000 16.9000000
% [115] 17.4500000 18.1000000 18.6666667 18.7166667 18.8000000 18.9333333
% [121] 20.1500000 20.6333333 20.7000000 20.8500000 21.3500000 23.4166667
% [127] 24.9333333 25.6500000 26.5666667 26.7666667 29.2166667 30.7000000
% [133] 30.9666667
% >

One limitation of using personal best times as indicators of
underlying athletic talent is that these times are potentially
confounded by the marathon courses at which they were set (some*sh
courses, like the Berlin marathon, are known for fast times) and race
conditions like weather.  In addition, personal best times may not
capture raceday idiosyncrasies that might affect individual runners.
With that in mind, Figure \ref{fig:45degreeplot_half} plots marathon
finishing times against half marathon splots.  This figure has a
regression line as before but no 45-degree line.

Figure \ref{fig:45degreeplot_half} shows that there was nothing
abnormal about the Hahner twins' overall finishing times, conditional
on their half marathon split times.  This is consistent with the
previous Figure \ref{fig:45degreeplot} and inconsistent with the
accusations against the Hahner twins, at least the part of the
accusation that focuses on their excessively slowing down.

If the Hahner twins did not slow down excessively, might they have
run somewhat strategically at the end of the race in order to generate
a simultaneous finish?  We now offer a visualization that speaks to
this question.

The personal best times of the Hahner twins were 115 seconds apart and
their official finishing times were separated by one second.  Is such
a 115 to one compression typical among pairs of runners or was it
unusual?  We might ask, are there other pairs of runners who had a
difference between personal best times of 115 seconds apart and, if
so, how close where their finishing times?

Of the 133 marathon finishers, there are $\binom{133}{2} = 8,778$
pairs of runners.  Of these and ignoring the Hahner twins, ten had
exactly a 115 second gap in personal best times.  Of these ten pairs,
differences in finishing times in seconds are as follows: 36, 93, 172,
319, 379, 459, 552, 671, 675, 739.  In other words, of all pairs of
runners in the Rio marathon who had a personal best difference that
was equivalent to the Hahner twins' difference, the twins had the
greatest compression based on finishing time.

% df2b[abs(df2b$pb_diff)==115,]
% # A tibble: 11 × 8
%                       name_i              name_j pb_diff result_diff
%                        <chr>               <chr>   <int>       <int>
% 1                Anna Hahner         Lisa Hahner    -115          -1
% 2                  Lily Luik Mariya Korobitskaya     115          36
% 3        Lisa Jane Weightman      Lanni Marchant    -115          93
% 4           Irina Smolnikova      Viviana Chávez     115        -172
% 5              Sonia Samuels    Helalia Johannes     115        -319
% 6  Jessica Draskau-Petersson       Gladys Tejeda     115         379
% 7        Maryna Damantsevich       Gladys Tejeda     115         459
% 8          Panayióta Vlaháki     Matea Matosevic    -115         552
% 9         Nataliya Lehonkova       Olha Kotovska     115         671
% 10               Anja Scherl      Desireé Linden     115         675
% 11              Rose Chelimo    Helalia Johannes    -115        -739
% # ... with 4 more variables: diff_in_diff <int>, consecutive <lgl>,
% #   twins <chr>, PERCENTILE <dbl>
% >

\begin{figure}[!ht]
 \caption{Differences in personal best times and finishing times}
 \label{fig:diffdiffscatter}
 \centering
 \begin{subfigure}{.5\textwidth}
 \includegraphics[width=\textwidth, keepaspectratio]{diff_in_diff_scatter_plot.pdf}
 \end{subfigure}
\end{figure}

We can generalize this result by looking at all pairs of runners in
the marathon.  For the 8,778 pairs of 133 finishers, Figure
\ref{fig:diffdiff} plots differences in finishing times against
differences in personal best times, and pairs of twins/triplets are
identified by the same color scheme we have used earlier.

Consider first the Hahner twins.  They are effectively located on the
horizontal axis (because their difference in finishing times is one
second).  However, there are many points about the Hahner's orange
dot, and this shows that, conditional on a 115 second difference in
personal best times,

In the
context of the marathon, that is, did Anna and Lisa Hahner run with an
eye on each other?

qq

@@

\begin{figure}[!ht]
  \caption{Relationship between Personal Best and Result  -- caption?}
  \label{fig:scatter}
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[width=\textwidth, keepaspectratio]{scatter_plot.pdf}
    \caption{Rio finishing times and personal best times}
    \label{fig:45degreeplot}
  \end{subfigure}
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[width=.975\textwidth, keepaspectratio]{studentized_residuals.pdf}
    \caption{Studentized residuals}
    \label{fig:studentizedresiduals}
  \end{subfigure}
\end{figure}

@@

% \begin{figure}[!ht]
%  \caption{Relationship between Difference in Personal Best and Difference in Result}
%  \label{fig:diffdiffscatter}
%  \centering
%  \begin{subfigure}{.5\textwidth}
%  \includegraphics[width=\textwidth, keepaspectratio]{diff_in_diff_scatter_plot.pdf}
%  \end{subfigure}
% \end{figure}
